---
title: "SitRep Production"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SitRep Production}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  root.dir = "C:/Users/lanak/Desktop/Sitrep"
)

library(SaviR)
library(tidyverse)
library(gt)

sunday_date <-  lubridate::floor_date(Sys.Date(), "week", week_start = 7)
```

## Data: The Frankenstein Dataset

onetable - stored metadata with iso2 and iso3 codes, country names, WHO/State/WorldBank regions, and UN population estimates
get_covid_df() - COVID dataframe from WHO and appended with select JHU obs
calc_add_risk() - Add risk matrix calculations to dataframe
get_vax() - Get vaccination data from OWID

This snippet produces the master dataframe with all the variables required for all of the sitrep graphics.

```{r}
df <- onetable %>%
  right_join(get_covid_df() %>% select(-who_region), by = c("iso2code" = "country_code")) %>%
  calc_add_risk() %>%
  left_join(get_vax(), by = c("id" = "iso_code", "date"))
```

## Epicurves

plot_epicurve(df, transparent = T)

The "plot_epicurve" function should take in the df with only WHO observations in order to match the same curve on the who.int dashboard. The individual epicurves for each region can be run simply by filtering for one region in the dataframe. The function will detect that the region column has only one region and will produce the epicurve accordingly.

Data prep: In this snippet, I first filter out any N/A WHO regions so the final plot won't have an extra unknown region, and I also filter just for WHO data (again in order to match who.int).

```{r}
# Global
plot_epicurve(filter(df, !is.na(who_region) & source == "WHO"), transparent = T) %>%
  ggsave(filename = "epi_curve_transparent_who.png", width = 13, height = 7, units = "in", bg = "transparent")

plot_epicurve(filter(df, !is.na(who_region)  & source == "WHO"), transparent = F)%>%
  ggsave(filename = "epi_curve_who.png", width = 13, height = 7, units = "in")

#Regions
for (r in unique(filter(df, !is.na(who_region))$who_region)){
  plot_epicurve(filter(df, !is.na(who_region) & source == "WHO" & who_region == r), transparent = F) %>%
    ggsave(filename = paste0("epi_curve_", Sys.Date(), "_", r, ".png"), width = 9, height = 2.5, units = "in", bg = "transparent")
}

```

## Burden Map

map_burden(df)

The "map_burden" function requires a result column where weekly average incidence is converted into factors using "cut". The dataframe normally used is WHO appended with JHU and filtered for the report date. The region maps can also be produced by simply filtering the input dataframe for the specific region. Again the function will detect the number of regions provided and will act accordingly.

```{r}
map_burden(df %>%
             filter(id != "CHN" | source != "WHO") %>%
             mutate(result = cut(week_case_incidence, breaks = c(-0.1, 1, 10, 25, Inf))) %>%
             filter(date == sunday_date)) %>%
  ggsave(filename = paste0("epi_curve_", Sys.Date(), "_", r, ".png"), width = 9, height = 2.5, units = "in", bg = "transparent")

for (r in unique(filter(df, !is.na(who_region))$who_region)){
  map_burden(df %>%
               filter(id != "CHN" | source != "WHO") %>%
               filter(date == sunday_date) %>%
               filter(who_region == r) %>%
               mutate(result = cut(week_case_incidence, breaks = c(-0.1, 1, 10, 25, Inf)))) %>%
    ggsave(filename = paste0("burdenmap_", Sys.Date(), "_", r, ".png"), width = 6, height = 6, units = "in", bg = "white")
}

```
## Trend Map

map_trend(df)

The "map_trend" function requires a result column where weekly percent change is converted into factors using "cut". The dataframe normally used is WHO appended with JHU and filtered for the report date. The region maps can also be produced by simply filtering the input dataframe for the specific region. Again the function will detect the number of regions provided and will act accordingly.

```{r}
map_trend(df %>%
            filter(date == sunday_date) %>%
            filter(id != "CHN" | source != "WHO") %>%
            mutate(result = cut(percent_change_case, breaks = c(-Inf, -50, 0, 50, 100, 200, Inf)))) %>%
  ggsave(filename = paste0("percent_change_who_", Sys.Date(), ".png"), width = 7.5, height = 4.2, units = "in", bg = "white")

for (r in unique(filter(df, !is.na(who_region))$who_region)){
  map_trend(df %>%
              filter(date == sunday_date) %>%
              filter(id != "CHN" | source != "WHO") %>%
              filter(who_region == r) %>%
              mutate(result = cut(percent_change_case, breaks = c(-Inf, -50, 0, 50, 100, 200, Inf)))) %>%
    ggsave(filename = paste0("percent_change_who_", Sys.Date(), "_", r, ".png"), width = 6, height = 6, units = "in", bg = "white")
}
```

## Vaccine Map (Global Only)

map_vaccinations(df, type = c("People", "Fully"))

```{r}
df %>% 
  filter(date <= sunday_date) %>%
  filter(id != "CHN" | source != "WHO") %>%
  mutate(result = cut(people_vaccinated_per_hundred, breaks = c(0, 1, 3, 10, 30, Inf))) %>%
  group_by(country) %>%
  filter(!is.na(result)) %>%
  filter(date == max(date)) %>%
  ungroup() %>%
  map_vaccinations(., type = "People") %>%
  ggsave(filename =  paste0("vaccine_map_", Sys.Date(), ".png"), width = 7.5, height = 4.2, units = "in", bg = "white")
```


## Top 10 Most Cases Table

```{r}
df %>%
  filter(date == sunday_date) %>%
  filter(id != "CHN" | source != "WHO") %>%
  select(country = who_country, value1 = week_case, value2 = percent_change_case) %>%
  arrange (desc(value1)) %>%
  head(10) %>%
  table_10mostcases(., run_date = format(sunday_date, "%B %d, %Y")) %>%
  gtsave(filename =  paste0("most_cases_table_t1", Sys.Date(), ".png"), vwidth=650)

for (r in unique(filter(df, !is.na(who_region))$who_region)){
  df %>%
    filter(date == sunday_date) %>%
    filter(id != "CHN" | source != "WHO") %>%
    filter(who_region == r) %>%
    select(country = who_country, value1 = week_case, value2 = percent_change_case) %>%
    arrange (desc(value1)) %>%
    head(10) %>%
    table_10mostcases(., type = r, run_date = format(sunday_date, "%B %d, %Y"))%>%
  gtsave(filename =  paste0('_wkcase_table_', Sys.Date(), "_", r, ".png"), vwidth = 650)
}
```

## Top 10 Incidence Table

```{r}
df %>%
  filter(date == sunday_date) %>%
  filter(id != "CHN" | source != "WHO") %>%
  select(country = who_country, value1 = week_case_incidence, value2 = percent_change_case) %>%
  arrange (desc(value1)) %>%
  head(10) %>%
  table_10incidence(., run_date = format(sunday_date, "%B %d, %Y")) %>%
  gtsave(filename =  paste0("incidence_table_t2", Sys.Date(), ".png"), vwidth = 650)

for (r in unique(filter(df, !is.na(who_region))$who_region)){
  df %>%
    filter(date == sunday_date) %>%
    filter(id != "CHN" | source != "WHO") %>%
    filter(who_region == r) %>%
    select(country = who_country, value1 = week_case_incidence, value2 = percent_change_case) %>%
    arrange (desc(value1)) %>%
    head(10) %>%
    table_10incidence(., type = r, run_date = format(sunday_date, "%B %d, %Y"))%>%
  gtsave(filename =  paste0('_wkcase_table2_', Sys.Date(), "_", r, ".png"), vwidth = 650)
}
```

## Top 10 Percent Change Table

```{r}
df %>%
  filter(date == sunday_date) %>%
  filter(id != "CHN" | source != "WHO") %>%
  filter(population >= 10000000) %>% #Exclude population less than 10M
  select(country = who_country, value1 = percent_change_case, value2 = percent_change4_case) %>%
  arrange (desc(value1)) %>%
  head(10) %>%
  table_10percentchange(., run_date = format(sunday_date, "%B %d, %Y")) %>%
  gtsave(filename =  paste0("pct_chng_table_output_who_t3", Sys.Date(), ".png"), vwidth=650)

for (r in unique(filter(df, !is.na(who_region))$who_region)){
  df %>%
    filter(date == sunday_date) %>%
    filter(id != "CHN" | source != "WHO") %>%
    filter(population >= 100000) %>% #Exclude population less than 100,000
    filter(who_region == r) %>%
    select(country = who_country, value1 = percent_change_case, value2 = percent_change4_case) %>%
    arrange (desc(value1)) %>%
    head(10) %>%
    table_10percentchange(., type = r, run_date = format(sunday_date, "%B %d, %Y"))%>%
  gtsave(filename =  paste0('_wkcase_table3_', Sys.Date(), "_", r, ".png"), vwidth = 650)
}
```

## Top 10 Vaccinations Table

table_10vaccinations(df, type = Global or Region, run_date)

The "table_10vaccinations" function takes looks for a country, value1 (people vaccinated per hundred), and a value2 (daily vaccinations per hundred) column. Note as vaccination reporting has gaps, the dataframe must be sliced by country and the most recent date with people_vaccinated_per_hundred value (if there is one). 

```{r}
df %>% 
  filter(date <= sunday_date) %>%
  filter(id != "CHN" | source != "WHO") %>%
  filter(population > 1000000) %>%
  group_by(country) %>%
  filter(!is.na(people_vaccinated_per_hundred)) %>%
  filter(date == max(date)) %>%
  ungroup() %>%
  select(country = who_country, value1 = people_vaccinated_per_hundred, value2 = daily_vaccinations_per_hundred) %>%
  arrange(desc(value1)) %>%
  head(10) %>%
  table_10vaccinations(., run_date = format(sunday_date, "%B %d, %Y")) %>%
  gtsave(filename =  paste0("top_10_vax_table", Sys.Date(), ".png"), vwidth = 2000)

for (r in unique(filter(df, !is.na(who_region))$who_region)){
  df %>%
    filter(date <= sunday_date) %>%
    filter(id != "CHN" | source != "WHO") %>%
    filter(population > 100000) %>% #Exclude population less than 100,000
    filter(who_region == r) %>%
    group_by(country) %>%
    filter(!is.na(people_vaccinated_per_hundred)) %>%
    arrange(date) %>%
    top_n(1, date) %>%
    distinct(id, .keep_all = T) %>%
    select(country = who_country, value1 = people_vaccinated_per_hundred, value2 = daily_vaccinations_per_hundred) %>%
    arrange(desc(value1)) %>%
    head(10) %>%
    table_10vaccinations(., type = r, run_date = format(sunday_date, "%B %d, %Y")) %>%
    gtsave(filename =  paste0("top_10_vax_table", Sys.Date(), "_", r, ".png"), vwidth = 700)
}
```

## Top 10 Fully Vaccinated Table

```{r}
table_10fullyvaccinations <- function(df, type = "Global", run_date = "Enter a date"){

  if(type == "Global"){
    title_label <-  gt::html(paste0("<b>Top 10 Countries/ Areas with Highest <br> Fully Vaccinated per 100 People", "</b>"))
    exclude_note <- "Countries with population size less than or equal to 1 million were excluded"
  } else {
    title_label  <- gt::html(paste0("<b>10 (", type, ") Countries/ Areas with Highest <br> Vaccination per 100 People", "</b>"))
    exclude_note <- "Countries with population size less than or equal to 100,000 were excluded"
  }

  gt::gt(df) %>%
    gt::tab_header(title    = title_label)%>%
    gt::fmt_number(columns  = c(value1),
                   decimals = 1) %>%
    gt::fmt_number(columns  = c(value2),
                   sep_mark = ",",
                   decimals = 2 ) %>%
    gt::data_color(columns  = c(value1),
                   colors   = scales::col_bin(palette = c("#CCECE6", "#AFDACB", "#92C8B1", "#75B696", "#57A37C", "#3A9161", "#1D7F47", "#006D2C"),
                                      bins    = c(0, 3, 10, 20, 30, 40, 60, 70, Inf))) %>%
    gt::cols_label(country  = "Country/ Area",
                   value1   = gt::html("People Fully Vaccinated <br> per 100 People"),
                   value2   = gt::html("Daily Vaccines <br> Administered <br> per 100 People")) %>%
    gt::cols_align("center") %>%
    gt::tab_options(column_labels.font.weight = "bold",
                table.font.size           = 20,
                table.font.weight         = "bold",
                footnotes.font.size       = pct(70),
                source_notes.font.size    = pct(70),
                source_notes.padding      = 0,
                footnotes.padding         = 0) %>%

    gt::tab_footnote(footnote  = "Number of people out of 100 who are fully vaccinated",
                     locations = cells_column_labels(columns = c(value1))) %>%
    gt::tab_footnote(footnote  = "Vaccine doses administered per day (7 day rolling average); does not represent number of people vaccinated",
                     locations = cells_column_labels(columns = c(value2))) %>%
    gt::tab_footnote(footnote  = exclude_note,
                     locations = cells_title()) %>%
    gt::tab_source_note(source_note = paste0("Data as of ", run_date))
}

df %>% 
  filter(date <= sunday_date) %>%
  filter(id != "CHN" | source != "WHO") %>%
  filter(population > 1000000) %>%
  group_by(country) %>%
  filter(!is.na(people_fully_vaccinated_per_hundred)) %>%
  filter(date == max(date)) %>%
  ungroup() %>%
  select(country = who_country, value1 = people_fully_vaccinated_per_hundred, value2 = daily_vaccinations_per_hundred) %>%
  arrange(desc(value1)) %>%
  head(10) %>%
  table_10fullyvaccinations(., run_date = format(sunday_date, "%B %d, %Y")) %>%
  gtsave(filename =  paste0("top_10_vax_table", Sys.Date(), ".png"), vwidth = 650)
```


## Risk Matrix (Global Only)

plot_riskmatrix(df)

Note for labeling purposes, the labels variable should be added to the dataframe specifying the top 10 countries for weekly average incidence, week case percent change, and weekly total cases. Exclusion criteria: Countries with population less than 10 million and countries with weekly cases totals less than 100.

```{r}
df %>%
  filter(date == sunday_date) %>%
  filter(id != "CHN" | source != "WHO") %>%
  filter(population >= 10000000) %>% #Exclude population less than 10M
  filter(week_case >= 100) %>%
  mutate(rank_inc = dense_rank(desc(week_case_incidence)),
         rank_per = dense_rank(desc(percent_change_case)),
         rank_cas = dense_rank(desc(week_case))) %>%
  mutate(labels   = case_when(rank_inc %in% 1:10 ~ country,
                              rank_per %in% 1:10 ~ country,
                              rank_cas %in% 1:10 ~ country)) %>%
  plot_riskmatrix() %>%
  ggsave(filename =  paste0('global_risk_matrix_who_', Sys.Date(), ".png"), width = 11, height = 7, units = "in")
           
```

## Vaccination Coverage (Global Only)

```{r}
df %>% 
  filter(date <= sunday_date) %>%
  filter(id != "CHN" | source != "WHO") %>%
  group_by(country, source) %>%
  arrange(desc(date)) %>%
  fill(total_vaccinations, .direction = "updown") %>%
  fill(people_vaccinated_per_hundred, .direction = "updown") %>%
  ungroup() %>%
  filter(date == max(date)) %>%
  group_by(who_region) %>%
  mutate(rank_people = dense_rank(-people_vaccinated_per_hundred),
         rank_total  = dense_rank(-total_vaccinations)) %>%
  mutate(labels      = case_when(rank_people %in% 1:3 ~ country,
                                 rank_total  %in% 1:3 ~ country)) %>%
  ungroup() %>%
  plot_vaxcoverage()
```

